---
title: "STA640 Homework 2"
author: "Fan Bu"
date: "9/12/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Problem 1
```{r, message=FALSE, warning=FALSE}
library(tableone)
library(survey)
library(reshape2)
library(ggplot2)
library(tidyverse)
data = read_delim('HW2_data.txt', delim=' ',
                  col_types = 'fnffffffnnnf')
```

### (a)
Estimate propensity score with logistic regression using all covariates as main effects (note that I've tried adding higher-order terms, but it tends to hinder algorithm convergence and doesn't improve balance by much). Then obtain the inverse probability weights.
```{r, message=FALSE}
# centering the continuous variables helps...
data[,c('i_age','com_t','pcs_sd','mcs_sd')] = scale(data[,c('i_age','com_t','pcs_sd','mcs_sd')],scale=FALSE)

ps.model = glm(pg ~ i_age + i_sex + i_race + i_educ + 
                 i_insu + i_drug + i_seve + com_t + 
                 pcs_sd + mcs_sd, 
               data=data, family=binomial('logit'))
```

```{r, echo=FALSE}
ps.model = readRDS('ps_model.rds')
```

```{r}
ps = ps.model$fitted.values %>% as.numeric()

data$ipw = 1
data$ipw[data$pg == 2] = 1/ps[data$pg == 2]
data$ipw[data$pg == 1] = 1/(1-ps[data$pg == 1])

data$ow = 1
data$ow[data$pg == 2] = 1 - ps[data$pg == 2]
data$ow[data$pg == 1] = ps[data$pg == 1]
```

Examine covariate balance. Here we have achieved good balance (ASD < 0.1) for some of the covariates, but for the others, there is still imbalance.

Note that there does exist a couple of subjects (subject 5 and 130) with extremely low ($<0.001$) or extremely high ($>0.999$) estimated propensity scores. And here I will exclude those extreme subjects. 
```{r}
## Weighted data
dataIPW <- svydesign(ids = ~ 1, 
                     data = data[ps > 0.001 & ps < 0.999,], 
                     weights = ~ ipw)

vars = names(data)[2:11]

## Construct a table that checks balance
tabWeighted <- svyCreateTableOne(vars = vars, strata = 'pg',
                                 data = dataIPW, test = FALSE)
## Show table with SMD
print(tabWeighted, smd = TRUE)
```


### (b)

Below is the histogram of estimated propensity scores by group.

```{r, echo=FALSE}
data$ps = ps

ggplot(data, aes(x=ps, fill=pg)) +
  geom_histogram(bins=30, position='dodge') +
  labs(fill='physician\ngroup', x='propensity score')
```

And also the Love plot showing ASD before and after matching.

```{r, echo=FALSE}
# nothing
tabUnmatched <- CreateTableOne(vars = vars, strata = "pg", data = data, test = FALSE)

## Weighted data (OW)
dataOW <- svydesign(ids = ~ 1, data = data, weights = ~ ow)

## Construct a table that checks balance
tabWeightedOW <- svyCreateTableOne(vars = vars, strata = 'pg',
                                 data = dataOW, test = FALSE)

```


```{r, echo=FALSE}
## Construct a data frame containing variable name and SMD from all methods
dataPlot <- data.frame(variable  = rownames(ExtractSmd(tabUnmatched)),
                       Unweighted = as.numeric(ExtractSmd(tabUnmatched)),
                       IPW  = as.numeric(ExtractSmd(tabWeighted)),
                       OW = as.numeric(ExtractSmd(tabWeightedOW)))

## Create long-format data for ggplot2
dataPlotMelt <- melt(data          = dataPlot,
                     id.vars       = c("variable"),
                     variable.name = "Method",
                     value.name    = "SMD")

## Order variable names by magnitude of SMD
varNames <- as.character(dataPlot$variable)[order(dataPlot$Unweighted)]

## Order factor levels in the same order
dataPlotMelt$variable <- factor(dataPlotMelt$variable,
                                levels = varNames)

## Plot using ggplot2
ggplot(data = dataPlotMelt,
       mapping = aes(x = variable, y = SMD, group = Method, color = Method)) +
    #geom_line() +
    geom_hline(yintercept = 0.1, color = "black", size = 0.3) +
    geom_point() +
    coord_flip() +
    theme_bw() + theme(legend.key = element_blank())
```


We can see that the original data (before weighting) do not have good balance on most of the covariates, in particular `race` and `comorbidity`, but there is some overlap between the two groups. After weighting, we achieve better balance in the covariates. 

### (c)

```{r get weighted average function, echo=FALSE}
weight_avg <- function(x, w, varname=NULL){
  if(is.numeric(x)){
    # continuous variable
    res = sum(w*x)/sum(w)
    names(res) = varname
    res
  }else{
    res = NULL
    names = NULL
    for(l in levels(x)){
      freq.l = sum((x==l) * w)/sum(w)
      res = c(res, freq.l)
      names = c(names, paste0(varname,'=',l))
    }
    names(res) = names
    res
  }
}

get_var_wa <- function(dat, varname){
  res = NULL
  
  x = dat[,varname,drop=TRUE]
  n = nrow(dat)
  
  # unweighted
  uni_W = rep(1,n)
  res = cbind(res, weight_avg(x[dat$pg==1], uni_W[dat$pg==1], varname))
  res = cbind(res, weight_avg(x[dat$pg==2], uni_W[dat$pg==2], varname))
  res = cbind(res, weight_avg(x, uni_W, varname))
  
  # IPW
  res = cbind(res, weight_avg(x, dat$ipw, varname))
  
  # OW
  res = cbind(res, weight_avg(x, dat$ow, varname))
  
  res
}

```


```{r construct the weighted average table, echo=FALSE}
tab = NULL

data_orig = read_delim('HW2_data.txt', delim=' ',
                  col_types = 'fnffffffnnnf')

# get original continuous columns
data_orig$ipw = data$ipw
data_orig$ow = data$ow
data_orig$ps = data$ps

for(v in vars){
  tab = rbind(tab, get_var_wa(data_orig, v))
}

tab = as.data.frame(tab)
names(tab) = c('PG 1 (unadj.)', 'PG 2 (unadj.)', 
               'Overall (unadj.)', 'IPW (adjusted)', 
               'OW (adjusted)')
knitr::kable(tab,digits=3)
```


<!-- |     |     |Unadjusted |       |    Adjusted |   | -->
<!-- |-----|------|-------|-------|-------|----------| -->
<!-- |     | **PG 1** | **PG 2** | **Overall** |  **IPW** |  **OW**  | -->
<!-- | Age (yrs) |  40.45 | 39.65 |        |  -->


